Natural selection in contemporary humans is driven by correlations with human capital

### 2021-04-13 why-natural-selection.Rmd
update paper
Include model in main paper

---
title: "Natural selection in contemporary humans is linked to income and substitution effects"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. Studying these can help us understand how the genetic architecture of health outcomes arise. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores associated with lower (higher) earnings, education and health are selected for (against). Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents (22+), or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market. We find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

These results show that selection effects are weaker, absent, or even reversed
among some subgroups of the population. One possible explanation for this is
given by the economic theory of fertility 
[@becker1960economic; @willis1973new;@becker1976child]. According to this theory,
increases in a person's wage affect their fertility via two opposing channels.
There is an *income effect* by which children become more affordable, like any
other good. There is also a *substitution effect*: since childrearing has a cost
in time, the opportunity cost of childrearing increases if one's market wage is
higher. The income effect would lead higher earners to have more children. The
substitution effect would lead higher earners to have fewer children. If so,
then genetic variants which affect earnings potential in the labour market may
cause opposing effects on fertility. The income effect will cause natural
selection in favour of earnings-increasing variants. The substitution effect
will do the reverse.

@@

>>>>

---
title: "Natural selection in contemporary humans is driven by correlations with human capital"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores associated with lower (higher) earnings, education and health are selected for (against). Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents (22+), or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
  - \usepackage{xr-hyper}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

These results show that selection effects are weaker, absent, or even reversed
among some subgroups of the population. One possible explanation for this is
given by the economic theory of fertility 
[@becker1960economic; @willis1973new;@becker1976child]. According to this theory,
increases in a person's wage affect their fertility via two opposing channels.
There is an *income effect* by which children become more affordable, like any
other good. There is also a *substitution effect*: since childrearing has a cost
in time, the opportunity cost of childrearing increases if one's market wage is
higher. The income effect would lead higher earners to have more children. The
substitution effect would lead higher earners to have fewer children. If so,
then genetic variants which affect earnings potential in the labour market may
cause opposing effects on fertility. The income effect will lead to natural
selection in favour of earnings-increasing variants. The substitution effect
will do the reverse.

To demonstrate this, we write down a simple model of 
fertility choices. $h$ is an individual's level of human capital, which 
(for now) we identify with his or her wage $W$. Raising a child takes time $b$. People maximize utility
from the number of children $N$, and income $Y\equiv(1-bN)W$. An
individual's payoff is

\[
U(N)=u(Y)+aN.
\]

Here $a$ captures the strength of preference for children. $u(\cdot)$
is concave and increasing. We treat $N$ as continuous, in line with
the literature: this can be thought of as the expected number of children
among similar people. The marginal benefit of an extra child is
$\frac{dU}{dN}=-bWu'(Y)+a$. The effect of an increase in wages (or human capital) 
on this marginal benefit is

\[
\frac{d^{2}U}{dNdh}=\underbrace{-bu'(Y)}_{\textrm{Substitution effect}}\underbrace{-bYu''(Y)}_{\textrm{Income effect}}.
\]

The *substitution effect* is negative and reflects that when
wages increase, time devoted to childcare costs more in foregone income.
The positive *income effect* reflects that when wage income is
higher, the marginal loss of wages from children is less painful.
Letting $u(\cdot)$ take the constant relative risk aversion (CRRA)
form $u(y)=\frac{y^{1-\sigma}-1}{1-\sigma}$, where $\sigma>0$ captures
risk aversion in income, we solve for optimal fertility $N^{*}$,
and examine the *fertility-human capital relationship*, $\frac{dN^{*}}{dh}$.
For moderate levels of risk aversion ($0.5<\sigma<1$):

\begin{enumerate}
\item The fertility-human capital relationship is negative: $\frac{dN^{*}}{dh}<0$.
\item The relationship is weaker (closer to zero) at higher wages and/or
levels of human capital.
\item The relationship is more negative when the time burden of children,
$b$, is larger.
\end{enumerate}

To examine education and fertility timing, we extend the model to
two periods. For convenience we ignore time discounting, and assume
that credit markets are imperfect so that agents cannot borrow. Write

\begin{equation}
U(N_{1},N_{2})=u(Y_{1})+u(Y_{2})+a(N_{1}+N_{2})\label{eq:U}
\end{equation}

Instead of identifying human capital $h$ with wages, we now allow
individuals to choose a level of education $s\in[0,1]$, which has
a time cost in period 1. Education increases period 2 wages, and is
complementary to human capital $h>0$. For period 2 wages we assume
the simple functional form $w(s,h)=sh$. We normalize period 1 wages
to 1. Maintaining the assumption of CRRA utility, we examine total
fertility $N^{*}\equiv N_{1}^{*}+N_{2}^{*}$. For $0.5<\sigma<1$,
facts 1-3 continue to hold (fact 2 in a neighborhood of $\sigma=1$).
In addition, for $\sigma$ in a neighborhood of 1:

\begin{enumerate}
\item [4.]\setcounter{enumi}{4}The fertility-human capital relationship
is weaker at higher levels of education $s$.
\item The relationship is weaker among those who start fertility in period
2 ($N_{1}^{*}=0$) than among those who start fertility in period
1 ($N_{1}^{*}>0$).
\end{enumerate}

Facts 1-5 match our empirical results. In this model, polygenic scores which
correlate with human capital would have a negative correlation with fertility at
low income and education levels, and a weaker or zero correlation at higher
levels (Figure \@ref(fig:plot-income-educ-level)). For scores which are highly
correlated with human capital, the correlation would also be weaker at higher
levels of the score itself, giving rise to diversifying selection (Appendix
Figure \@ref(fig:plot-purifying)). The correlation would be more negative when
the time burden of children $b$ is high, e.g. for single parents (Figure
\@ref(fig:plot-n-partners)). Lastly, the correlation would be weaker among those
starting fertility later (Figure \@ref(fig:plot-age-flb)).

An alternative theory is that polygenic scores correlate with the
motivation to have children (parameter $a$ in the model). However, this theory
would not explain why our results are weaker at higher incomes and education levels.
Indeed, in the one-period model, while $\frac{dN^{*}}{da}>0$, this
relationship actually becomes stronger at higher wages.

The above does not provide a complete theory of fertility. Rather,
it shows that a relatively simple economic model can explain many
of our results. In the appendix, we discuss other models of fertility
from the economic literature, as well as causal evidence for the relationship
between human capital, income and fertility.

@@

\clearpage
\FloatBarrier

# Proofs

```{r include-model-appendix, child="model-appendix.Rmd"}
